---
title: "Student performance"
author: "William Pattison"
date: "27/07/2020"
output: html_document
---

# Load data and libraries
```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(corrplot)
library(GGally)
library(DataExplorer)
library(ggplot2)
library(caret)
library(pls)
library(Cubist)
students_performance <- read_csv("StudentsPerformance.csv")
```

# EDA

```{r EDA}
str(students_performance)
```

## Change the types of some of the variables in the dataset as they are incorrect. (make new, fixed dataset)

```{r}
c_students_performance <- students_performance
c_students_performance$`test preparation course` <- as.factor(c_students_performance$`test preparation course`)
c_students_performance$lunch <- as.factor(c_students_performance$lunch)
c_students_performance$`parental level of education` <- as.factor(c_students_performance$`parental level of education`)
c_students_performance$`race/ethnicity` <- as.factor(c_students_performance$`race/ethnicity`)
c_students_performance$gender <- as.factor(c_students_performance$gender)
```

## Recheck the dataset
```{r}
summary(c_students_performance)
```

## visualisations

### Distribtuion of features

```{r}
plot_bar(c_students_performance)
plot_histogram(c_students_performance)
```
Various visualisations about the data

### Correlation

```{r}
corr <- cor(c_students_performance[6:8])
corrplot.mixed(corr)
```

### math score by gender
```{r}
p <- ggplot(c_students_performance, aes(`math score`)) + geom_histogram(binwidth=5, color="gray", aes(fill=gender))
p <- p + xlab("Math Score") + ylab("Gender") + ggtitle("Math Score by gender")
p
```

### writing scores by gender
```{r}
p <- ggplot(c_students_performance, aes(`writing score`)) + geom_histogram(binwidth=5, color="gray", aes(fill=gender))
p <- p + xlab("Writing Score") + ylab("Gender") + ggtitle("Writing Score by gender")
p
```

### reading scores by gender
```{r}
p <- ggplot(c_students_performance, aes(`reading score`)) + geom_histogram(binwidth=5, color="gray", aes(fill=gender))
p <- p + xlab("Reading Score") + ylab("Gender") + ggtitle("Reading Scores by gender")
p
```

### Math score by preparation
```{r}
p <- ggplot(c_students_performance, aes(`test preparation course`, `math score`))
p <- p + geom_boxplot()
p <- p + ggtitle("Math score by preparation")
p <- p + xlab("Preparation") + ylab("Math Score")
p
```

### Writing score by preparation

```{r}
p <- ggplot(c_students_performance, aes(`test preparation course`, `writing score`))
p <- p + geom_boxplot()
p <- p + ggtitle("Writing score by preparation")
p <- p + xlab("Preparation") + ylab("Writing Score")
p
```

### Reading score by preparation

```{r}
p <- ggplot(c_students_performance, aes(`test preparation course`, `reading score`))
p <- p + geom_boxplot()
p <- p + ggtitle("Reading score by preparation")
p <- p + xlab("Preparation") + ylab("Reading Score")
p
```

The same process above could be done with other factor variables such as education level

# Modeling

## Data split (test/train)
```{r}
set.seed(222)

train_index <- createDataPartition(c_students_performance$`writing score`, p = .75,
                                   list = F)
training <- c_students_performance[train_index,]
testing <- c_students_performance[-train_index,]
```

## Training and tuning

```{r}
fit_control <- trainControl(method = "boot", number = 10, repeats = NA, search = "random")

lm1 <- train(`writing score` ~ ., data = training, method = "lm", 
             metric = "RMSE", trControl = fit_control)
pls1 <- train(`writing score` ~ ., data = training, method = "pls", 
              metric = "RMSE", trControl = fit_control)
cubist1 <- train(`writing score` ~ ., data = training, method = "cubist", 
              metric = "RMSE", trControl = fit_control)
null1 <- train(`writing score` ~ ., data = training, method = "null", 
              metric = "RMSE", trControl = fit_control)

final_pls1 <- tolerance(pls1$results, metric = "RMSE", 
                         tol = 2, maximize = F)  

final_cubist1 <- tolerance(cubist1$results, metric = "RMSE", 
                         tol = 2, maximize = F)  
```

## Modelling results

### lm
```{r}
print(lm1)
```

### pls
```{r}
print(pls1)
plot(pls1)
```


### cubist

```{r}
print(cubist1)
plot(cubist1)
```

### null

```{r}
print(null1)
```


### comparison of models training

```{r}
resamps <- resamples(list(lm = lm1,
                          cubist = cubist1,
                          pls = pls1,
                          null = null1))
dotplot(resamps, metric = "RMSE")

resamps_no_null <- resamples(list(lm = lm1,
                          cubist = cubist1,
                          pls = pls1))
#better look at the results
dotplot(resamps_no_null, metric = "RMSE")
```


## test data

```{r}
lm1_predictions <- predict(lm1, newdata = testing)
d1 <- data.frame(testing$`writing score`, lm1_predictions)
colnames(d1) <- c("obs", "pred")

pls1_predictions <- predict(pls1, newdata = testing)
d2 <- data.frame(testing$`writing score`, pls1_predictions)
colnames(d2) <- c("obs", "pred")

cubist1_predictions <- predict(cubist1, newdata = testing)
d3 <- data.frame(testing$`writing score`, cubist1_predictions)
colnames(d3) <- c("obs", "pred")
```


### plots

```{r}
lm1_predictions <- predict(lm1, newdata = testing)
d1 <- data.frame(testing$`writing score`, lm1_predictions)
colnames(d1) <- c("obs", "pred")
plot(d1, main = "Predicted versus Observed for test data (lm)")
abline(a = 0, b = 1, col = c("blue"), lty = c(2), lwd = c(3))

pls1_predictions <- predict(pls1, newdata = testing)
d2 <- data.frame(testing$`writing score`, pls1_predictions)
colnames(d2) <- c("obs", "pred")
plot(d2, main = "Predicted versus Observed for test data (pls)")
abline(a = 0, b = 1, col = c("blue"), lty = c(2), lwd = c(3))
postResample(pred = pls1_predictions, obs = testing$`writing score`)

cubist1_predictions <- predict(cubist1, newdata = testing)
d3 <- data.frame(testing$`writing score`, cubist1_predictions)
colnames(d3) <- c("obs", "pred")
plot(d3, main = "Predicted versus Observed for test data (cubist)")
abline(a = 0, b = 1, col = c("blue"), lty = c(2), lwd = c(3))
```

### metrics

```{r}
print("lm")
postResample(pred = lm1_predictions, obs = testing$`writing score`)
print("pls")
postResample(pred = pls1_predictions, obs = testing$`writing score`)
print("cubist")
postResample(pred = cubist1_predictions, obs = testing$`writing score`)
```